---
title: "Visualization of gene expression with Nebulosa"
author: "Jose Alquicira-Hernandez"
package: Nebulosa
output: 
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Visualization of gene expression with Nebulosa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

# Overview

`Nebulosa` is an R package to visualize expression data from single cells based
on kernel density estimation. Due to the sparsity observed in single-cell data 
(e.g. RNA-seq, ATAC-seq), the visualization of cell features is frequently 
affected and unclear, especially when it is overlayed with clustering
to annotate cell types. `Nebulosa` integrates the "cell density" from a low
dimensional space along with the expression values of a cell feature (e.g. 
gene, peak) to draw more informed conclusions about the identity and the
dynamics of cell populations.

# Import libraries
  
For this vignette, let's use `Nebulosa` with the `Seurat` package. 
First, we'll do a brief/standard data processing.

```{r import_libraries}
library("Nebulosa")
library("Seurat")
```

# Data pre-processing


Let's download a dataset of 3k PBMCs (available from 10X Genomics). This same 
dataset is commonly used in Seurat vignettes. The code below will download, 
store, and uncompress the data in a temporary directory.

```{r download_and_untar_file}
download.file(
  "https://s3-us-west-2.amazonaws.com/10x.files/samples/cell/pbmc3k/pbmc3k_filtered_gene_bc_matrices.tar.gz",
  file.path(tempdir(), "pbmc3k.tar.gz")
)

untar(file.path(tempdir(), "pbmc3k.tar.gz"),
  exdir = file.path(
    tempdir(),
    "pbmc3k"
  )
)
```


Then, we can read the gene expression matrix using the `Read10X` from `Seurat`

```{r read_data}
data <- Read10X(data.dir = file.path(
  tempdir(),
  "pbmc3k",
  "filtered_gene_bc_matrices",
  "hg19"
))
```

Let's create an object with features being expressed in at least 3 cells and 
cells expressing at least 200 genes.

```{r create_seurat_object}
pbmc <- CreateSeuratObject(
  counts = data,
  project = "pbmc3k",
  min.cells = 3,
  min.features = 200
)
```

Remove outlier cells based on the number of genes being expressed in each cell 
(below 2500 genes) and expression of mitochondrial genes (below 5%).

```{r qc}
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")
pbmc <- subset(pbmc, subset = nFeature_RNA < 2500 & percent.mt < 5)
```

## Data normalization

Let's use `SCTransform` to stabilize the variance of the data by regressing out 
the effect of the sequencing depth from each cell.

```{r norm, message=FALSE, warning=FALSE}
pbmc <- SCTransform(pbmc, verbose = FALSE)
```

## Dimensionality reduction

Once the data is normalized and scaled, we can run a 
_Principal Component Analysis_  (PCA) first to reduce the dimensions of our 
data from 26286 features to 50 principal components. To visualize the principal 
components, we can run a
_Uniform Manifold Approximation and Projection for Dimension Reduction_ (UMAP) 
using the first 30 principal components to obtain a two-dimentional space.

```{r dim_red, message=FALSE, warning=FALSE}
pbmc <- RunPCA(pbmc)
pbmc <- RunUMAP(pbmc, dims = 1:30)
```

## Clustering

To assess cell similarity, let's cluster the data by constructing a 
_Shared Nearest Neighbor _(SNN) _Graph_ using the first 30 principal components 
and applying the _Louvain algorithm_.

```{r clustering, message=FALSE, warning=FALSE}
pbmc <- FindNeighbors(pbmc, dims = 1:30)
pbmc <- FindClusters(pbmc)
```


# Visualize data with `Nebulosa`

The main function from `Nebulosa` is the `plot_density`. For usability, it 
resembles the `FeaturePlot` function from `Seurat`. 

Let's plot the kernel density estimate for `CD4` as follows

```{r plot_cd4}
plot_density(pbmc, "CD4")
```

For comparison, let's also plot a standard scatterplot using `Seurat`

```{r cd4_comparison}
FeaturePlot(pbmc, "CD4")
FeaturePlot(pbmc, "CD4", order = TRUE)
```

By smoothing the data, `Nebulosa` allows a better visualization of the global 
expression of CD4 in myeloid and CD4+ T cells. Notice that the "random" 
expression of CD4 in other areas of the plot is removed as the expression of 
this gene is not supported by many cells in those areas. Furthermore, CD4+ 
cells appear to show considerable dropout rate.

Let's plot the expression of CD4 with `Nebulosa` next to the clustering results

```{r cd4_and_clustering}
DimPlot(pbmc, label = TRUE, repel = TRUE)
```

We can now easily identify that clusters `0` and `2` correspond to CD4+ T cells
if we plot CD3D too.

```{r plot_cd3d}
plot_density(pbmc, "CD3D")
```

# Multi-feature visualization

Characterize cell populations usually relies in more than a single marker. 
Nebulosa allows the visualization of the joint density of from multiple 
features in a single plot.

## Identifying Naive CD8+ T cells

Users familiarized with PBMC datasets may know that CD8+ CCR7+ cells usually 
cluster next to CD4+ CCR7+ and separate from the rest of CD8+ cells. Let's aim 
to identify Naive CD8+ T cells. To do so, we can just add another gene to the 
vector containing the features to visualize. 

```{r fig.height=14}
p3 <- plot_density(pbmc, c("CD8A", "CCR7"))
p3 + plot_layout(ncol = 1)
```

When compared to the clustering results, we can easily identify that Naive 
CD8+ T cells correspond to cluster `8`. 


`Nebulosa` returns the density estimates for each gene along with the joint 
density across all provided genes. By setting `combine = FALSE`, we can obtain 
a list of ggplot objects where the last plot corresponds to the joint density 
estimate.


```{r}
p_list <- plot_density(pbmc, c("CD8A", "CCR7"), combine = FALSE)
p_list[[length(p_list)]]
```

## Identifying Naive CD4+ T cells

Likewise, the identification of Naive CD4+ T cells becomes straightforward by 
combining `CD4` and `CCR7`:

```{r fig.height=14}
p4 <- plot_density(pbmc, c("CD4", "CCR7"))
p4 + plot_layout(ncol = 1)
```

Notice that these cells are mainly constrained to cluster `0`

```{r fig.height=10}
p4[[3]] / DimPlot(pbmc, label = TRUE, repel = TRUE)
```
